---
title: 'Capital Bikeshare Dataset Analysis'
author: "Michael Alpas, Mohit Singh and Upendra Yadav"
date: "August 07, 2020"
output:
  html_document: 
    theme: readable
    toc: yes  
  pdf_document: default
urlcolor: cyan
---
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80)
```

## Introduction

---

### Overview

For this project, we would like to simulate that Team Outlier is a team of Data Scientists that Capital Bikeshare company hired to come up with data driven answers to help them with their decisions.

The following business questions will be guided by the two-year historical log corresponding to years 2011 and 2012.

 * Capital Bikeshare would like to have a targeted marketing strategy. The Team Outlier need to determine the best season to have marketing promotional offers to increase the number of customers.
 * The company also need to reduce operational expenses by looking at the manpower needed. Team Outlier will predict the number of people needed during Holidays vs Regular day.

### Description of Dataset

---

### Dataset Source

---

## Methods

---

### Targeted Marketing Strategy Analysis
#### Loading the Dataset

```{r}
library(readr)
day_data = read.csv("dataset/day.csv")
# knitr::kable(head(day_data)[, 1:15])
head(day_data)

hour_data = read.csv("dataset/hour.csv")
# knitr::kable(head(hour_data)[, 1:15])
head(hour_data)
```

### Filter the data based on season
```{r}
# spring filtered data
spring_data = subset(day_data, day_data$season == 1)

# summer filtered data
summer_data = subset(day_data, day_data$season == 2)

# fall filtered data
fall_data   = subset(day_data, day_data$season == 3)

# winter filtered data
winter_data = subset(day_data, day_data$season == 4)
```

### Remove Possible NA

```{r}
# remove NA
spring_data = na.omit(spring_data)
summer_data = na.omit(summer_data)
summer_data = na.omit(fall_data)
winter_data = na.omit(winter_data)
```

### Quick Comparison

let's take a quick look with the filtered data.

```{r}
# plot to see the difference 
par(mfrow=c(2,2))
hist(spring_data$cnt, col = "dodgerblue",border = "darkorange", xlab = "count", main = "Spring Daily Rental Bike Summary")
hist(summer_data$cnt, col = "dodgerblue",border = "darkorange", xlab = "count", main = "Summer Daily Rental Bike Summary")
hist(summer_data$cnt, col = "dodgerblue",border = "darkorange", xlab = "count", main = "Fall Daily Rental Bike Summary")
hist(winter_data$cnt, col = "dodgerblue",border = "darkorange", xlab = "count", main = "Winter Daily Rental Bike Summary")

```

### Operational Expenses Analysis

---

## Results

---

## Discussion

---

## Appendix

---

